
---
title: "Lab 1: Airbnbs in NYC"
format: html
editor: source
---


## Airbnb in NYC (or your city)

Airbnb has had a disruptive effect on the hotel, rental home, and 
vacation industry throughout the world. The success of Airbnb has not come without controversies, with critics arguing that Airbnb has adverse impacts on housing and rental
prices and also on the daily lives of people living in neighborhoods where Airbnb is popular.
This controversy has been particularly intense in NYC, where the debate been Airbnb proponents
and detractors eventually led to the city imposing strong restrictions on the use of Airbnb. If you find this issue interesting and want to go deeper, there is the potential for
an interesting project that brings in hotels (which have interesting regulations in NYC), hotel price data, and rental data and looks at these things together.

Because Airbnb listings are available online through their website and app, it is possible
for us to acquire and visualize the impacts of Airbnb on different cities, including 
New York City. This is possible through the work of an organization called 
[inside airbnb](https://insideairbnb.com/get-the-data/)

## Github Instructions

Before we introduce the data and the main assignment, let's begin with a few key 
steps to configure the file and create a github repository for your first assignment. This is optional
but I think it is a good idea to start getting familiar with github tools. These steps come from
the happygitwithr tutorial. 

-   Start a new github repository in your account, clone it to your computer (using RStudio to start a new project from a repository or any other way)
-   Update the YAML, changing the author name to your name, and **Render** the document.
-   Commit your changes with a meaningful commit message.
-   Push your changes to GitHub.
-   Go to your repo on GitHub and confirm that your changes are visible in your Qmd **and** md files. If anything is missing, commit and push again.

## Packages

We'll use the **tidyverse** package for much of the data wrangling and visualisation, and
the **ggridges** package to make a ridge plot in the last exercise. You may need to install
**ggridges** if you haven't already, you can do that using:

```{r}
#| eval: false

install.packages("ggridges")

```

Then make sure to load both packages:
```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(ggridges)
```

## Data

*The data for this assignment can be found on my github page [by clicking here and downloading nycbnb.csv](https://github.com/georgehagstrom/DATA607/tree/main/website/data/nycbnb.csv)*

If you
are adventurous and want to perform this assignment for a different city you can choose one 
from [inside airbnb)](https://insideairbnb.com/get-the-data/). If you go that route, make sure to download the file listings.csv.gz for the city you selected (gz is an archive format which you should be able to expand), and you will only need to keep the following columns for this analysis: 

```{r}
#| label: select
#| eval: false
  nycbnb = nycbnb |> 
    select(
      id, 
      price, 
      neighbourhood_cleansed,
      neighbourhood_group_cleansed,
      accommodates, 
      bathrooms, 
      bedrooms, 
      beds, 
      review_scores_rating, 
      number_of_reviews, 
      listing_url )
```

You will also need to do some post-processing, including changing the price column from a string to a numerical variable. If you decide to go this custom route let me know and make sure to share your data, but I recommend sticking with the data I provided.

You can read the data into R using the command:

```{r}
#| label: loadData
#| warning: false
#| include: true


nycbnb = read_csv("/home/georgehagstrom/work/Teaching/DATA608/DataStory4/nyc_airbnb_listings.csv")


```


where you should replace `/home/georgehagstrom/work/Teaching/DATA607/website/data/nycbnb.csv"nycbnb.csv` with the local path to your file. 

Important note: It is generally not wise to include datasets in github repositories, especially
if they are large and can change frequently. 

You can view the dataset as a spreadsheet using the `View()` function.
Note that you should not put this function in your R Markdown document, but instead type it directly in the Console, as it pops open a new window (and the concept of popping open a window in a static document doesn't really make sense...).
When you run this in the console, you'll see the following **data viewer** window pop up.

# Exercises
                                                                                              
**Problem 1.** How many observations (rows) does the dataset have? Instead of hard coding the number in your answer, use inline code. 
                                                                                               
**Problem 2.** Run `View(nycbnb)` in your Console to view the data in the data viewer. What does each row in the dataset represent?
                                                                                                                                         
Each column represents a variable. We can get a list of the variables in the data frame using the `names()` function.
                                                                                         
```{r}
#| label: namesCommand
#| warning: false


names(nycbnb)

```


You can find descriptions of each of the variables in the help file for the dataset, which
you can find online at the inside airbnb [data dictionary](https://docs.google.com/spreadsheets/d/1iWCNJcSutYqpULSQHlNyGInUvHg2BoUGoNRIGa6Szc4/edit?usp=sharing)
                                                                                               
**Problem 3.**  Pick one of the five boroughs of NYC (Manhattan, Queens, Brooklyn, the Bronx, or Staten Island), and create a faceted histogram where each facet represents a neighborhood in your chosen borough and displays the distribution of Airbnb prices in that neighborhood. Think critically about whether it makes more sense to stack the facets on top of each other in a column, lay them out in a row, or wrap them around. Along with your visualization, include your reasoning for the layout you chose for your facets.

R

```{r}
#| label: manhattan-price-histogram
#| fig-width: 10
#| fig-height: 8

library(tidyverse)
# Make sure the nyc_airbnb dataset is loaded first

nyc_airbnb = nycbnb

nyc_airbnb |>
  filter(borough == "Manhattan", price < 1000) |>
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 50, fill = "#FF5A5F", color = "white") +
  facet_wrap(~ neighborhood) +
  labs(
    title = "Distribution of Airbnb Prices by Neighborhood in Manhattan",
    subtitle = "For listings with a nightly price under $1000",
    x = "Price per Night ($)",
    y = "Number of Listings"
  ) +
  theme_bw()
```

```{r}
#| label: manhattan-price-histogram-improved
#| fig-width: 10
#| fig-height: 12

nyc_airbnb |>
  filter(borough == "Manhattan", price < 1000) |>
  ggplot(aes(x = price)) +
  geom_histogram(aes(y = after_stat(density)),binwidth = 50, fill = "#FF5A5F", color = "white") +
  # 1. Control the layout with fewer columns
  facet_wrap(~ neighborhood, ncol = 5) +
  labs(
    title = "Distribution of Airbnb Prices by Neighborhood in Manhattan",
    subtitle = "For listings with a nightly price under $1000",
    x = "Price per Night ($)",
    y = "Number of Listings"
  ) +
  theme_bw() +
  # 2. Adjust text sizes to be less obtrusive
  theme(
    strip.text = element_text(size = 8),      # Smaller neighborhood labels
    axis.title = element_text(size = 10),     # Slightly smaller axis titles
    axis.text = element_text(size = 7)        # Smaller numbers on the axes
  )

```

**Problem 4.**  Use a single pipeline to identity the neighborhoods city-wide with the top five median listing prices
that have a minimum of 50 listings. Then, in another pipeline filter the data for these five neighborhoods and make ridge plots of the distributions of listing prices in these five neighborhoods. In a third pipeline calculate the minimum, mean, median, standard deviation, IQR, and maximum listing price in each of these neighborhoods. Use the visualization and the summary statistics to describe the distribution of listing prices in the neighborhoods. (Your answer will include three pipelines, one of which ends in a visualization, and a narrative.)

```{r}

#| label: find-top-5-neighborhoods

library(tidyverse)
# Ensure the nyc_airbnb dataset is loaded
# nyc_airbnb <- read_csv("nyc_airbnb_listings.csv")

# Create a vector of the top 5 neighborhood names
top_5_neighbourhoods <- nyc_airbnb |>
  group_by(neighborhood) |>
  summarise(
    median_price = median(price, na.rm = TRUE),
    listing_count = n()
  ) |>
  filter(listing_count >= 50) |>
  arrange(desc(median_price)) |>
  slice_head(n = 5) |>
  pull(neighborhood)

# Print the names to see which they are
top_5_neighbourhoods



```


```{r}
#| label: ridge-plot-top-5
#| fig-width: 8
#| fig-height: 6

library(ggridges)

# Filter for the top 5 neighborhoods and prices < $1500, then plot
nyc_airbnb |>
  filter(neighborhood %in% top_5_neighbourhoods, price < 1500) |>
  ggplot(aes(x = price,
             y = fct_reorder(neighborhood, price, .fun = median),
             fill = neighborhood)) +
  geom_density_ridges(alpha = 0.8, scale = 0.9) +
  labs(
    title = "Price Distribution in NYC's Most Expensive Neighborhoods",
    subtitle = "For neighborhoods with at least 50 listings",
    x = "Price per Night ($)",
    y = "Neighborhood"
  ) +
  theme_minimal() +
  theme(legend.position = "none")




```

```{r}
#| label: summary-stats-top-5

nyc_airbnb |>
  filter(neighborhood %in% top_5_neighbourhoods) |>
  group_by(neighborhood) |>
  summarise(
    min_price = min(price, na.rm = TRUE),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE),
    st_dev = sd(price, na.rm = TRUE),
    iqr = IQR(price, na.rm = TRUE),
    max_price = max(price, na.rm = TRUE)
  ) |>
  arrange(desc(median_price))



```



**Problem 5.**  Create a visualization that will help you compare the distribution of review scores (`review_scores_rating`) across neighborhoods. You get to decide what type of visualization to create and which
neighborhoods are most interesting to you, and there is more than one correct answer! In your answer, include a brief interpretation of how Airbnb guests rate properties in general and how the neighborhoods compare to each other in terms of their ratings.
    
